Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-05-08 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
update
this is an alternative version/introduction. I have comment the previous section -introduction and intro to viscous instab- and I have created a new introduction that includes both of them while i tried to decrease the size.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Introduction}\label{section:intro}
Numerical investigation of multiphase flows in porous media have attracted the attention of the scientific community over the past 40 years. Characterisation of such fluid flows serves as the foundation of reservoir engineering and groundwater studies \citep{white_1981}. %Oil and gas migration in reservoirs and transport of contaminants in groundwater are some of important physical problems in which porous media flows play a crucial role. 
%Reservoir engineers and hydro-geologists are dealing with complex geometries, irregular boundaries and heterogeneous geological formations.
Underground coal gasification is another important field of interest and, more recently, due to the role of green house gases (GHG) emissions in the global climate change, several research work have focused on CO$_{2}$ migration and trapping mechanisms in carbon capture utilisation and storage (CCUS) operations \citep{spycher_2003, self_2012, jiang_2011}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\blue{\subsection{Alternative Introduction}}
\blue{- - - - - - - - - - - - - - - - - - - - - - - - - - - -}

@@

\medskip
Description of physics and mechanisms of multiphase porous media flows were reported by \citet{wooding_1976} with focus on capillary pressure and flow regimes. A comprehensive review of force balances at the interface between immiscible fluids and resulting mechanisms for flow instabilities can be found in \citet{homsy_1987}. Flow instabilities can be classified as macroscopic and microscopic, the former is due to imposed boundary conditions, whereas the later is associated to local phenomena at the fluids interface \citep[\eg Kelvin-Helmholtz and Saffman-Taylor instabilities,][]{saffman_1959}. This work focuses on flow instabilities in two-phase systems due to viscous and stress forces by neglecting gravitational effects (often referred as viscous instabilities or viscous fingering, Section \ref{section:ViscousInstabilities}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Brief Summary of Viscous Instabilities}\label{section:ViscousInstabilities}
The study of viscous flow instabilities (\ie fingering) is particularly important in oil exploration due to heterogeneities (\ie natural fractures, permeability and/or porosity characteristics in different zones) of geological formations. A major problem associated with water-flooding processes is the early water-breakthrough caused by high-permeability layers and unfavorable mobility ratios. Water breakthrough and volumetric sweep efficiency (ratio of volumes between the recovered and the injected fluids) are the main determinants of the productive life of a reservoir \citep{riaz_2004, tavassoli_2015}. During immiscible CO$_{2}$-flooding $\left(\text{\ie in CO}_{2}\text{ enhanced oil recovery, CO}_{2}\text{-EOR, processes}\right)$, viscosity of supercritical CO$_{2}$ is lower than crude oil, thus viscous fingering and/or channelling may occur.

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\begin{comment}
\section{Introduction}\label{section:intro}
Numerical investigation of multiphase flows in porous media have attracted the attention of the scientific community over the past 40 years. Characterisation of such fluid flows serves as the foundation of reservoir engineering and groundwater studies \citep{white_1981}. %Oil and gas migration in reservoirs and transport of contaminants in groundwater are some of important physical problems in which porous media flows play a crucial role. 
%Reservoir engineers and hydro-geologists are dealing with complex geometries, irregular boundaries and heterogeneous geological formations.
Underground coal gasification is another important field of interest and, more recently, due to the role of green house gases (GHG) emissions in the global climate change, several research work have focused on CO$_{2}$ migration and trapping mechanisms in carbon capture utilisation and storage (CCUS) operations \citep{spycher_2003, self_2012, jiang_2011}.

@@

\medskip
A brief description of the numerical formulation and viscous fluid instabilities are introduced in Sections~\ref{equations_scheme} and~\ref{section:ViscousInstabilities}, respectively. Model set up and results including initial model-benchmark are presented in Section~\ref{section:results}. Impact of mobility ratio on the fingers formation is also included in this section. Finally, concluding remarks are presented in Section~\ref{Section:Conclusion}.
\end{comment}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\blue{\section{Alternative Introduction}}
\blue{- - - - - - - - - - - - - - - - - - - - - - - - - - - -}

@@

\medskip
Description of physics and mechanisms of multiphase porous media flows were reported by \citet{wooding_1976} with focus on capillary pressure and flow regimes. A comprehensive review of force balances at the interface between immiscible fluids and resulting mechanisms for flow instabilities can be found in \citet{homsy_1987}. Flow instabilities can be classified as macroscopic and microscopic, the former is due to imposed boundary conditions, whereas the later is associated to local phenomena at the fluids interface \citep[\eg Kelvin-Helmholtz and Saffman-Taylor instabilities,][]{saffman_1959}. This work focuses on flow instabilities in two-phase systems due to viscous and stress forces by neglecting gravitational effects (often referred as viscous instabilities or viscous fingering). %Section \ref{section:ViscousInstabilities}).

\medskip
In oil and gas reservoir exploration, viscous and density instabilities are relatively common during water-injection processes. As water and oil interacts, the interface between these fluids moves creating an uneven or fingered flow profile (see Fig.~\ref{fig:simple_case}). Viscous fingering results in inefficient flow sweeping which can bypass significant quantities of recoverable oil and may lead to early breakthrough of water into neighbour production wells. This problem is associated with water-flooding processes and caused by high-permeability layers and unfavorable mobility ratios. Water breakthrough and volumetric sweep efficiency (ratio of volumes between the recovered and the injected fluids) are the main determinants of the productive life of a reservoir \citep{riaz_2004, tavassoli_2015}. During immiscible CO$_{2}$-flooding $\left(\text{\ie in CO}_{2}\text{ enhanced oil recovery, CO}_{2}\text{-EOR, processes}\right)$, viscosity of supercritical CO$_{2}$ is lower than crude oil, thus viscous fingering and/or channelling may occur.

\medskip
Viscous instabilities can be found across several disciplines and scales and are mainly controlled by the mobility ratio (MR) between displacing and displaced fluids. Other conditions that may also influence the severity of viscous fingering are: heterogeneity (\ie wide spatial porosity and/or permeability distribution), gravitational forces, anisotropic dispersion, non-monotonic viscosity profile etc \citep{budek_2017,nicolaides_2015}. \citet{muskat_1934} investigated fluid flow in Hele-Shaw cells, \ie parallel flat plates separated by an infinitesimal gap, and the impact on the capillary number ($N_{c}$),

\begin{equation}
N_{c} = \frac{\mu U}{\gamma}.
\label{eq:capillary_number}
\end{equation}

\medskip
In homogeneous domains, fingers start to develop when the surface tension acting on the interface between the fluids exert an opposite force towards the change of shape of the interface. The interface becomes unstable and collapses, taking a curved shape \citep{homsy_1987, jackson_2017}. In heterogeneous domains, such instability may be triggered by permeability differences across regions as demonstrated in Section~\ref{section:results_homo_hete}. The higher the velocity of the low viscosity fluid, the less wide (tip-splitting behaviour) the finger is. Pressure differences acting on the interface produces a net pressure force,

\begin{equation} 
\Delta P= - \gamma \nabla\cdot\hat{n}. 
\label{eq:pressure_dif} 
\end{equation}

\noindent where subscripts $w$ and $nw$ stand for wetting and non-wetting phases, respectively. $\mathcal{K}^{\circ}_{rw}$ and $\mathcal{K}^{\circ}_{rnw}$ are end-point relative permeability to wetting and non-wetting phases, $S_{w,irr}$ and $S_{nw,r}$ are irreducible wetting and residual non-wetting phase saturations, respectively. Exponents $n_{w}$ and $n_{nw}$ are both set to 2. The MR contains the phase saturation (see Eqns.~\ref{eq:MR}-\ref{Eqn:CoreyBrooks2}) while it is clear that during fluid displacement phase saturation, $S_{w}\left(x_{i},t\right)$ and $S_{nw}\left(x_{i},t\right)$, will change in time and space (Eqn.~\ref{eqn:saturation_eqn}). Therefore, with no lack of generality, the MR can be reduced to the viscosity ratio (VR),

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{comment} 
\section{Brief Summary of Viscous Instabilities}\label{section:ViscousInstabilities}
The study of viscous flow instabilities (\ie fingering) is particularly important in oil exploration due to heterogeneities (\ie natural fractures, permeability and/or porosity characteristics in different zones) of geological formations. A major problem associated with water-flooding processes is the early water-breakthrough caused by high-permeability layers and unfavorable mobility ratios. Water breakthrough and volumetric sweep efficiency (ratio of volumes between the recovered and the injected fluids) are the main determinants of the productive life of a reservoir \citep{riaz_2004, tavassoli_2015}. During immiscible CO$_{2}$-flooding $\left(\text{\ie in CO}_{2}\text{ enhanced oil recovery, CO}_{2}\text{-EOR, processes}\right)$, viscosity of supercritical CO$_{2}$ is lower than crude oil, thus viscous fingering and/or channelling may occur.

@@

%\pagebreak
%\clearpage
\end{comment}
